Tail associations and their impact on extinction risk

### 2019-11-03 MainText.Rmd
a bit of progress on the abstract

\noindent \textbf{ABSTRACT} 
<!--200 words limit : J Am Nat-->

<!--***DAN: Dan has not edited this--> 
With global climate change, extreme climatic events (ECEs) become more frequent. These ECEs could pose a severe impact on the resistance or resilience of a population and/ or community level through life-history traits of the species. Therefore predicting the impact of extreme events on population extinction risk is necessary. We here introduce the potential use of "tail association" property of two correlated ecological variables in modeling the extinction risk of an age-structured single patch model. We also extend this idea to density-dependent metapopulation models. For metapopulation models, we show extinction risk not only depends on the nature of the tail association of spatially correlated noise but is also influenced by the inherent dynamics of the population-model. To elaborate, we demonstrate left tail dependent noise across spatially correlated patches cause higher extinction risk than the right tail dependent noise when the model show undercompensatory dynamics. But in the overcompensatory regime, the trend is the opposite. Our study paves the path for future research considering this tail association instead of a simple correlation in population viability analysis.

\textbf{\textit{Keywords:}} extinction risk, extreme events, tail association, density-dependent model

@@

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
for both Q1 and Q2 instead of attempting mathematical 
analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms, outlined above, which suggest that
asymmetries of tail association may make a difference for extinction risks. 
By so doing we hope to open these potentially practically important questions to 
wider examination other researchers. Some material in the Discussion 
outlines some possible next steps.
<!--***DAN: make sure you do this-->
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

>>>>

\noindent \textbf{ABSTRACT} 
<!--350 words limit : Ecosphere-->

\noindent Extreme climatic events are becoming more frequent and more intense
due to climate change. Furthermore, there is reason to believe extreme 
climatic events may modify "tail associations" between values of an environmental 
variable measured in different locations, i.e., associations that 
occur principally between values in the left or right tails of the distributions 
of values of the variable that can occur in the individual locations. 
Two positively associated variables can be "left-tail associated" (i.e., more 
correlated when they take low than high values) or "right-tail associated"
(more correlated when they take high than low values) even with the same
overall correlation coefficient in both cases. We tested whether the nature of 
spatial tail associations of environmental variables can influence metapopulation 
extinction risk. For instance, if low values of a variable are bad for local
populations, one may expect that left-tail associations increase
metapopulation extinction risks because then environmental catastrophes are
spatially synchronized, presumably reducing the potential for rescue effects.
We also tested, in the context of (non-spatial) stage-structured matrix models, 
whether tail associations between values of stage-specific vital rates may
influence extinction risk.

Tail associations and their asymmetries are taken into account in assessing risks
in finance and other fields, but to our knowledge our study is one of the first
to consider how tail associations influence extinction risk.

With global climate change, extreme climatic events (ECEs) become more frequent. These ECEs could pose a severe impact on the resistance or resilience of a population and/ or community level through life-history traits of the species. Therefore predicting the impact of extreme events on population extinction risk is necessary. We here introduce the potential use of "tail association" property of two correlated ecological variables in modeling the extinction risk of an age-structured single patch model. We also extend this idea to density-dependent metapopulation models. For metapopulation models, we show extinction risk not only depends on the nature of the tail association of spatially correlated noise but is also influenced by the inherent dynamics of the population-model. To elaborate, we demonstrate left tail dependent noise across spatially correlated patches cause higher extinction risk than the right tail dependent noise when the model show undercompensatory dynamics. But in the overcompensatory regime, the trend is the opposite. Our study paves the path for future research considering this tail association instead of a simple correlation in population viability analysis.

\noindent \textbf{\textit{Keywords:}} extinction risk, extreme events, tail association, density-dependent model

@@

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
for both Q1 and Q2 instead of attempting mathematical 
analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms, outlined above, which suggest that
asymmetries of tail association may make a difference for extinction risks. 
By so doing we hope to open these potentially practically important questions to 
wider examination by other researchers. Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.
